The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-05-31 doc/SA-Hermaphrodites-wDemography.tex
Functions for  sex-specific dominance figs
This commit adds some new functions to plot supplementary figures
illustrating how the model behaves udner sex-specific dominance.

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the intrinsic population growth rate remains positive. In addition, we analyze the effects of inbreeding depression manifesting in different life-history stages, and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results indicate that as population intrinsic growth rates approach one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M. guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}). To take the population dynamical consequences of SA selection into account, a model is required that links changes in genotype distributions to population dynamics. The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few papers since then explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model of \citep{deVriesCaswell2019a,deVriesCaswell2019b} to include both dioecious and hermaphroditic species in order to study the parameter conditions under which balanced SA polymorphisms are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

Simultaneous hermaphrodites express both male and female sex functions (i.e., can transmit genes to the next generation via both sperm and ovules), and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to male sex function with a prime.

@@

Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. \textcolor{red}{In the \hl{Online Supplement}, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}). Do we???}

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For simplicity, we present results for scenarios of equal dominance in the main text (i.e., $h_f = h_m = h$). Specifically, we examine scenarios of ({\itshape i}) additive SA fitness effects ($h = 1/2$); and ({\itshape ii}) dominance reversals, where the deleterious fitness effect of each SA allele is partially recessive in each sex ($h = 1/4$). These scenarios are of particular interest because additive effects are commonly observed for small-effect alleles or quantitative traits \citep{AgrawalWhitlock2011}, while dominance reversals are predicted under fitness landscape models provided the population is not too far from the phenotypic optimum \citep{Manna2011, ConnallonClark2014}. \textcolor{red}{We explore the consequences of sex-specific dominance, which may evolve under asymmetric SA selection \citep{SpencerPriest2016} in \hl{Appendix X}).}

For a given parameter set, we numerically evaluate the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrate inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds

For analyses of the demographic effects of inbreeding depression, we make two additional simplifying assumptions: First, we assume that if inbreeding depression is caused primarily by recessive deleterious mutations (as suggested by empirical data), it should covary negatively with the population selfing rate due to purging (\citealt{Charlesworth2009}; though we note that other processes could give rise to this pattern, e.g., \citealt{CrnokrakBarrett2002, Charlesworth2009,HedrickGarcia-Dorado2016}). Following \citet{Olito2019}, we incorporate such negative covariance by constraining the inbreeding depression terms in the model ($\delta$ and $\delta_i$, where $i \in \{j,a,\gamma\}$) to follow a simple declining function of the selfing rate: $\delta = \delta^{\ast} (1 - b (1 - L))$, where $\delta^{\ast}$ is the hypothetical severity of inbreeding depression for a completely outcrossing population, $b$ is a shape parameter determining how far $\delta$ will decline under complete selfing (when $C = 1$), and $L$ describes the expected deleterious mutation load as a function of the selfing rate $C$. The function $L$ includes an additional shape parameter, $a$, which determines the curvature of the overall function for $\delta$ (see Appendix E in \citealt{Olito2019} for additional details). We set $\delta^{\ast} = 0.8$, $b = 0.5$, and $a = 0.2$ for all analyses, values chosen to be consistent with empirical estimates of inbreeding depression (e.g., fig.~2 in \citealt{HusbandSchemske1996}). Second, to keep our analyses tractable we explore the effects of individual inbreeding depression terms in isolation. That is, we assume that only one of the $\delta$ and $\delta_i$ terms (where $i \in \{j,a,\gamma\}$) can be non-zero at a time.

@@

However, a key prediction from the evolutionary demographic model is that large fractions of SA polymorphic parameter space can be demographically inviable (fig.~\ref{fig:extThresholds}). The location of the extinction threshold, where the population intrinsic growth rate $\lambda = 1$, is primarily determined by the fertility parameter ($f$) but is also influenced by the population selfing rate ($C$) and dominance of the SA alleles ($h$). When $f$ is large, the population can experience stronger selection on fertility through both male and female fitness components before extinction occurs (the extinction threshold shifts to higher values of $s_f$ and $s_m$).

@@

The point in the life-cycle where inbreeding depression manifests can influence the threshold selfing rate at which demographically viable polymorphic parameter space crashes. Our results indicate that population viability was most sensitive to inbreeding depression affecting juvenile survival rates ($\delta_j$; fig.~\ref{fig:deltaPolySpace}, dark blue points), while early-acting inbreeding depression ($\delta$, ovule abortion shortly after fertilization) had a similar effect on population viability as late-acting inbreeding depression affecting adult survial ($\delta_a$) and juvenile-to-adult transition rates ($\delta_{\gamma}$). Inbreeding effects on juvenile survival had the strongest effect on population viability because on average individuals will spend multiple time steps in the juvenile stage before they mature. At each time step in the juvenile stage, individuals have a probability $\gamma$ to mature and a probability $\sigma_j(1-\delta_j)$ to survive. Inbreeding depression at the juvenile stage therefore makes it harder to survive long enough to mature. Although early-acting inbreeding depression ($\delta$) actually manifests earlier in the life-cycle than juvenile survival, it acts only once by influencing the total number of self-fertilized zygotes that become juveniles.

@@

Interestingly, a polymorphic chromosomal inversion (inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. inv6 segregates at moderate frequency (about $8\%$), and carriers suffer $\approx 30 \%$ loss in pollen viability, but also increased flower production. The genetic basis of these effects remain unclear, but the net result is a "supergene" with strong fitness effects on both female and male fertility that segregates as a single diallelic locus. We estimated selection coefficients for the effect of inv6 on pollen production (i.e., taking into account the simultaneous effect on flower number) and ovule production from the data reported in \citet{LeeKelly2015} under the relatively conservative assumptions that pollen/ovule production is proportional to flower number, and  additive SA fitness effects (dominance coefficients could not be estimated from field data; see Appendix \hl{C} for details of estimating selection coefficients). Under these assumptions, the average selection coefficient for inv6 on flower production across two years in the field was $(s_f,s_m) = (0.3,0.31)$. As a final proof-of-concept for our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, inv6 appears to fall in demographically viable SA polymorphic parameter space (see Figure \ref{fig:mimulusFig}), as might be expected given its observed frequency in the Iron Mountain population.

@@

% Figure 4
\begin{figure}[htbp]
 \centering
 \includegraphics[scale=0.75]{../output/figs/mimulusinv6FigNewSel.pdf}
 \caption{\footnotesize{Illustration of model predictions using empirically estimated demographic rates for {\itshape M.~guttatus}. Results are shown for two hypothetical populations using demographic rates for locally adapted (Eagle Meadows; panel A), and non-local (Low-Elevation Perennial; panel B) populations reported by \citet{PetersonEtAl2016}. Invasion conditions and extinction thresholds were calculated using the conservative assumption of additive fitness effects in both sexes ($h_f = h_m = 0.5$) for two parameter conditions: obligate outcrossing (solid black lines)  and partial selfing with inbreeding depression using the highest field-estimate of selfing, and inbreeding depression parameters ($\delta_i$) calculated for the Iron Mountain population of \citet{Willis1993} (dashed lines). The location of inv6 is also shown on both plots, using selection coefficients calculated from field estimates of male and female fitness components from \citet{LeeKelly2015} under the relatively conservative assumption of additive fitness effects in both sexes, where $(s_f,\,s_m) = (0.30,\, 0.31)$. Note that in panel A the invasion conditions for partial selfing with inbreeding depression (dotted lines) are nearly indistinguishable from those for obligate outcrossing (solid lines).}} 
 \label{fig:mimulusFig}
 \end{figure}

@@

Classic population genetics theory predicts that sexually antagonistic selection is unlikely to maintain genetic variation except under narrow conditions, with polymorphism requiring either finely balanced or unusually strong selection, or partially recessive fitness effects through each sex \citep{Kidwell1977,Pamilo1979,Prout2000, ConnallonClark2014}. Extensions of the theory have identified numerous ways in which the conditions for polymorphism become more permissive in both dioecious and hermaphroditic organisms, including genetic linkage of SA loci, the evolutioin of sex-specific dominance, population subdivision, and life-cycle complexity \citep[e.g.,][]{Patten2010,JordanConnallon2014,JordanCharlesworth2012,SpencerPriest2016, ConnallonSharmaOlito2019, Olito-etal-2018}. However, by ignoring the demographic consequences of SA genetic variation, these population genetic models have missed the possibility that SA polymorphisms may not be viable under realistic parameter conditions, and therefore unlikely to be observed in natural populations. By linking the individual-level fitness consequences of SA selection to population level consequences, our theoretical framework provides several key insights into the processes shaping SA genetic variation in natural populations.

The first and central finding of our study is that when intrinsic population growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. Since growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exception of recently introduced, invasive populations), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. Moreover, these findings compliment recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be high \citep{BergerEtAl2014, Connallon2015}.

We also find that demographically viable parameter space is often biased towards alleles with stronger selection through the female than male sex function. That is, given natural variation in population growth rates, the most demographically viable (and therefore observable in natural populations) outcomes of SA selection are either fixation of a female-beneficial allele, or polymorphisms involving low-frequency female-deleterious alleles. This key prediction is supported by a series of experimental results in seed beetles ({\itshape Callosobruchus maculatus}) which show that male-beneficial SA genotypes are less likely to contribute positively to population growth rates, and are more susceptible to extinction under environmental stress or inbreeding \citep{BergerEtAl2014, berger2016intralocus, grieshop2017male}. An interesting corrollary is that, since strong SA fitness effects, especially those affecting female fertility, often lead to extinction, the SA genetic variation we are able to observe in natural populations may often be under weak selection, and therefore strongly susceptible to genetic drift whether or not selection favours the fixation of one allele or balancing selection \citep{ConnallonClark2012}.

In hermaphroditic populations, self-fertilization can alleviate the demographic costs of balanced SA polymorphisms under some conditions, however, the concommitant effects of inbreeding depression generally exacerbate them in populations with mixed-mating systems. This prediction is in stark contrast to previous population genetics models of SA selection in hermaphrodites, where the sole effect of inbreeding depression is to reduce the population effective selfing rate through the loss of selfed-zygotes, thereby expanding polymorphic parameter space \citep{JordanConnallon2014, Olito2017}. In our model, this reduction in effective selfing is accompanied by significant mortality due to inbreeding depression (whether early- or late-acting), which can quickly tip partially-selfing populations over the brink to extinction. Beyond the maintenance of SA polymorphisms, this finding underscores a simple but important point that is often overlooked in studies of the evolution of self-fertilization and selfing-syndromes, which tend to emphasize the coevolution of the deleterious mutation load and mating system \citep[e.g.,][]{Charlesworth1987, LandeSchemske1985, Goodwillie2005}: populations with high fertility rates are better able to sustain positive growth rates in the face of severe demographic costs of inbreeding. This suggests that traits related to female fecundity, such as ovule and flower production, in ancestral populations likely play an important role in determining the distribution of transitions to self-fertilization among hermaphroditic taxa as well as a variety of ecological correlates of selfing and mixed-mating \citep{Goodwillie2005,Igic2006,Grossenbacher2015}.

@@

Like most demographic matrix models, ours also assumes female demographic dominance, where population growth rates are determined entirely by female fecundity \citep{pollard1975mathematical,Caswell2001,iannelli2005gender}. Yet, many hermaphroditic populations experience limitation of reproductive success due to both quantity and quality of male gametes \citep[e.g.,][]{Yund2000,AizenHarder2007, Harder2016}. Explicitly modeling male gamete production is a natural extension to our modeling framework, and would enable us to analyze how tension between the demographic consequences of SA fitness variation through both female and male function alters or abolishes asymmetries in the extinction thresholds and polymorphic parameter space \citep[e.g.,][]{Tazzyman2015}.  Interestingly, different forms of self-fertilization can aid in reproductive assurance, and should therefore have demographic consequences -- for example, under pollen limitation and delayed selfing (where only ovules that fail to receive outcross pollen are selfed), the selfing rate will be a function of genotype frequencies because these will directly influence pollen production \citep{HarderBarrett2006}.
,
Another major simplifying assumption in our model was that how individuals themselves were produced (i.e., by selfing or outcrossing) affects the level of inbreeding depression they suffer. In reality, the history of consecutive generations of inbreeding in each individual's lineage will influence the severity of inbreeding depression they experience, particularly when inbreeding depression is caused primarily by recessive deleterious mutations. It would be an interesting and feasible extension of our modeling framework to expand the individual state space to include selfing cohorts (i.e., first generation selfing, second generation selfing, etc.), as in the models of \citet{kelly1999response,kelly2007mutation}, thereby enabling a more biologically plausible approach to modelling self-fertilization.

@@

As described in the main text, we incorporated empirical estimates of the population selfing rate and inbreeding depression for two intensively studied populations in the Cascade Mountains of Oregon into our parameterized model (Iron Mountain and Cone Peak; \citealt{Willis1993,Willis1999a,Willis1999b}). Using the data provided in Table.~2 of \citet{Willis1993}, we calculated the proportional decrease due to inbreeding depression in three fitness components that were also included in vital rate estimates of \citet{PetersonEtAl2016}: seed germination rate ($G$), flower number ($F$), and overwinter survival ($S$). In each case, we calculated the inbreeding depression terms as $\delta_i = 1 - w_{\text{inbred}}/w_{\text{outcross}}$. To incorporate these terms into our evolutionary demographic model, we multiplied each of the three terms $G$, $F$, and $S$ in Eq(\ref{eq:AtildeMimulus}) by a corresponding proportional inbreeding depression term $(1 - \delta_i)$, where $i \{\in G,F,S\}$. The largest field-estimated selfing rate for this same Iron Mountain population was $C = 0.29$ \citep{Willis1993}, which we use in our calculations for fig.~\ref{fig:mimulusFig}.

>>>>

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying parameter conditions under which SA polymorphism is maintained {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant (and often infinite) size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}). To take into account the consequences of population dynamics under SA selection, a model is required that links changes in genotype distributions to population dynamics. The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few papers since then explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model of \citep{deVriesCaswell2019a,deVriesCaswell2019b} to include both dioecious and hermaphroditic species in order to study the parameter conditions under which different outcomes of SA selection, including fixation of male- or female-beneficial alleles as well as balanced polymorphisms, are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

Simultaneous hermaphrodites express both male and female sex functions (i.e., can transmit genes to the next generation via both sperm/pollen and eggs/ovules), and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to male sex function with a prime.

@@

Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. In the \hl{Online Supplement}, we present a more general model of genotype-specific self-fertilization rates (after \citealt{JordanConnallon2014}).

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. We present scenarios of equal dominance in the main text (i.e., $h_f = h_m = h$). Specifically, we examine scenarios of ({\itshape i}) additive SA fitness effects ($h = 1/2$), which are commonly observed for small-effect alleles or quantitative traits \citep{AgrawalWhitlock2011}; and ({\itshape ii}) dominance reversals, where the deleterious fitness effect of each SA allele is partially recessive in each sex ($h = 1/4$), which are predicted under fitness landscape models provided the population is not too far from the phenotypic optimum \citep{Manna2011, ConnallonClark2014}. We briefly explore the consequences of sex-specific dominance, which may evolve under asymmetric SA selection \citep{SpencerPriest2016} in \hl{Appendix X}).

For a given parameter set, we numerically evaluate the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrate inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds.

For analyses of the demographic effects of inbreeding depression, we make two additional simplifying assumptions: First, we assume that if inbreeding depression is caused primarily by recessive deleterious mutations (as suggested by empirical data), it should covary negatively with the population selfing rate due to purging, provided the population selfing rate has been relatively constant in recent evolutionary time (\citealt{Charlesworth2009}; though we note that other processes could give rise to this pattern, e.g., \citealt{CrnokrakBarrett2002, Charlesworth2009,HedrickGarcia-Dorado2016}). Following \citet{Olito2019}, we incorporate such negative covariance by constraining the inbreeding depression terms in the model ($\delta$ and $\delta_i$, where $i \in \{j,a,\gamma\}$) to follow a simple declining function of the selfing rate: $\delta = \delta^{\ast} (1 - b (1 - L))$, where $\delta^{\ast}$ is the hypothetical severity of inbreeding depression for a completely outcrossing population, $b$ is a shape parameter determining how far $\delta$ will decline under complete selfing (when $C = 1$), and $L$ describes the expected deleterious mutation load as a function of the selfing rate $C$. The function $L$ includes an additional shape parameter, $a$, which determines the curvature of the overall function for $\delta$ (see Appendix E in \citealt{Olito2019} for additional details). We set $\delta^{\ast} = 0.8$, $b = 0.5$, and $a = 0.2$ for all analyses, values chosen to be consistent with empirical estimates of inbreeding depression (e.g., fig.~2 in \citealt{HusbandSchemske1996}). Second, to keep our analyses tractable we explore the effects of individual inbreeding depression terms in isolation. That is, we assume that only one of the $\delta$ and $\delta_i$ terms (where $i \in \{j,a,\gamma\}$) can be non-zero at a time.

@@

However, a key prediction from the evolutionary demographic model is that large fractions of SA polymorphic parameter space can be demographically inviable (fig.~\ref{fig:extThresholds}). The location of the extinction threshold, where the population intrinsic growth rate $\lambda = 1$, is primarily determined by the fertility parameter ($f$) but is also influenced by the population selfing rate ($C$) and dominance of the SA alleles ($h$). As $f$ increases, the population can survive in the face of stronger fertility selection through both male and female sex-functions before extinction occurs (the extinction threshold shifts to higher values of $s_f$ and $s_m$).

@@

The point in the life-cycle where inbreeding depression manifests can influence the threshold selfing rate at which demographically viable polymorphic parameter space crashes. Our results indicate that population viability was most sensitive to inbreeding depression affecting juvenile survival rates ($\delta_j$; fig.~\ref{fig:deltaPolySpace}, dark blue points), while early-acting inbreeding depression ($\delta$, ovule abortion shortly after fertilization) had a similar effect on population viability as late-acting inbreeding depression affecting adult survial ($\delta_a$) and juvenile-to-adult transition rates ($\delta_{\gamma}$). Inbreeding effects on juvenile survival had the strongest effect on population viability because on average individuals will spend multiple time steps in the juvenile stage before they mature. At each time, juveniles have a probability $\gamma$ to mature and a probability $\sigma_j(1-\delta_j)$ to survive. Inbreeding depression at the juvenile stage therefore makes it harder to survive long enough to mature. Although early-acting inbreeding depression ($\delta$) actually manifests earlier in the life-cycle than juvenile survival, it acts only once by influencing the total number of self-fertilized zygotes that become juveniles. Note, however, that the relative strength of inbreeding depression at different stages of the life history will depend on the interaction between the various inbreeding depression parameters, which we have precluded from our analyses.

@@

Interestingly, a polymorphic chromosomal inversion (inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. inv6 segregates at moderate frequency (about $8\%$), and carriers suffer $\approx 30 \%$ loss in pollen viability, but also increased flower production. The genetic basis of these effects remain unclear -- though they likely polygenic -- but the net result is a "supergene" with strong fitness effects on both female and male fertility that segregates as a single diallelic locus. We estimated selection coefficients for the effect of inv6 on pollen production (i.e., taking into account the simultaneous effect on flower number) and ovule production from the data reported in \citet{LeeKelly2015} under the relatively conservative assumptions that pollen/ovule production is proportional to flower number, and  additive SA fitness effects (dominance coefficients could not be estimated from field data; see Appendix \hl{C} for details of estimating selection coefficients). Under these assumptions, the average selection coefficient for inv6 on flower production across two years in the field was $(s_f,s_m) = (0.3,0.31)$. As a final proof-of-concept for our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, inv6 appears to fall in demographically viable SA polymorphic parameter space (see Figure \ref{fig:mimulusFig}), as might be expected given its observed frequency in the Iron Mountain population.

@@

% Figure 4
\begin{figure}[htbp]
 \centering
 \includegraphics[scale=0.75]{../output/figs/mimulusinv6FigNewSel.pdf}
 \caption{\footnotesize{Illustration of model predictions using empirically estimated demographic rates for {\itshape M.~guttatus}. Results are shown for two hypothetical populations using demographic rates for locally adapted (Eagle Meadows; panel A), and non-local (Low-Elevation Perennial; panel B) populations reported by \citet{PetersonEtAl2016}. Invasion conditions and extinction thresholds were calculated using the conservative assumption of additive fitness effects in both sexes ($h_f = h_m = 0.5$) for two parameter conditions: obligate outcrossing (solid black lines, dark red shaded region)  and partial selfing with inbreeding depression using the highest field-estimate of selfing, and inbreeding depression parameters ($\delta_i$) calculated for the Iron Mountain population of \citet{Willis1993} (dashed lines, light red shaded region). The location of inv6 is also shown on both plots, using selection coefficients calculated from field estimates of male and female fitness components from \citet{LeeKelly2015} under the relatively conservative assumption of additive fitness effects in both sexes, where $(s_f,\,s_m) = (0.30,\, 0.31)$. Note that in panel A the invasion conditions for partial selfing with inbreeding depression (dotted lines) are nearly indistinguishable from those for obligate outcrossing (solid lines).}} 
 \label{fig:mimulusFig}
 \end{figure}

@@

Classic population genetics theory predicts that sexually antagonistic selection is unlikely to maintain genetic variation except under narrow conditions, with polymorphism requiring either finely balanced or unusually strong selection, or partially recessive fitness effects through each sex \citep{Kidwell1977,Pamilo1979,Prout2000, ConnallonClark2014}. Extensions of the theory have identified numerous ways in which the conditions for polymorphism become more permissive in both dioecious and hermaphroditic organisms, including genetic linkage of SA loci, the evolution of sex-specific dominance, population subdivision, and life-cycle complexity \citep[e.g.,][]{Patten2010,JordanConnallon2014,JordanCharlesworth2012,SpencerPriest2016, ConnallonSharmaOlito2019, Olito-etal-2018}. However, by ignoring the demographic consequences of SA genetic variation, these population genetic models have missed the possibility that SA polymorphisms may not be viable under realistic parameter conditions, and therefore unlikely to be observed in natural populations. By linking the individual-level fitness consequences of SA selection to population level consequences, our theoretical framework provides several key insights into the processes shaping SA genetic variation in natural populations.

The first and central finding of our study is that when intrinsic population growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. Since growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exception of recently introduced, invasive populations), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. Moreover, these findings complement recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be high \citep{BergerEtAl2014, Connallon2015}.

We also find that demographically viable parameter space is often biased towards alleles with stronger selection through the female than male sex function. That is, given natural variation in population growth rates, the most demographically viable (and therefore observable in natural populations) outcomes of SA selection are either fixation of a female-beneficial allele, or polymorphisms involving low-frequency female-deleterious alleles. This key prediction is supported by a series of experimental results in seed beetles ({\itshape Callosobruchus maculatus}) which show that male-beneficial SA genotypes are less likely to contribute positively to population growth rates, and are more susceptible to extinction under environmental stress or inbreeding \citep{BergerEtAl2014, berger2016intralocus, grieshop2017male}. \textcolor{red}{Jessica's comment Re: sex-limited exp. evol.}. An interesting corrollary of our findings is that, since strong SA fitness effects often lead to extinction, the observable SA genetic variation in natural populations may often be under weak selection, and therefore strongly susceptible to genetic drift whether or not selection favours the fixation of one allele or balancing selection \citep{ConnallonClark2012}.

In hermaphroditic populations, self-fertilization can alleviate the demographic costs of balanced SA polymorphisms under some conditions, however, the concommitant effects of inbreeding depression generally exacerbate them in populations with mixed-mating systems. This prediction is in stark contrast to previous population genetics models of SA selection in hermaphrodites, where the sole effect of inbreeding depression is to reduce the population effective selfing rate through the loss of selfed-zygotes, thereby expanding polymorphic parameter space \citep{JordanConnallon2014, Olito2017}. In our model, this reduction in effective selfing is accompanied by significant mortality due to inbreeding depression (whether early- or late-acting), which can quickly tip partially-selfing populations over the brink to extinction. Beyond the maintenance of SA polymorphisms, this finding underscores a simple but important point that is often overlooked in studies of the evolution of self-fertilization and selfing-syndromes, which tend to emphasize the coevolution of the deleterious mutation load and mating system \citep[e.g.,][]{Charlesworth1987, LandeSchemske1985, Goodwillie2005}: highly fertile populations can better afford the severe demographic costs of inbreeding. This suggests that traits related to female fecundity, such as ovule and flower production, may strongly influence the distribution of successful transitions to self-fertilization among hermaphroditic taxa, as well as a variety of ecological correlates of selfing and mixed-mating \citep{Goodwillie2005, Igic2006, Grossenbacher2015}.

@@

Like most demographic matrix models, ours also assumes female demographic dominance, where population growth rates are determined entirely by female fecundity \citep{pollard1975mathematical,Caswell2001,iannelli2005gender}. Yet, many hermaphroditic populations experience limitation of reproductive success due to both quantity and quality of male gametes \citep[e.g.,][]{Yund2000,AizenHarder2007, Harder2016}. Explicitly modeling male gamete production is a natural extension to our modeling framework, and would enable us to analyze how tension between the demographic consequences of SA fitness variation through both female and male function alters or abolishes asymmetries in the extinction thresholds and polymorphic parameter space \citep[e.g.,][]{Tazzyman2015}.  Interestingly, different forms of self-fertilization can aid in reproductive assurance, and should therefore have demographic consequences -- for example, under pollen limitation and delayed selfing (where only ovules that fail to receive outcross pollen are selfed), the selfing rate will be a function of genotype frequencies because these will directly influence pollen production \citep{HarderBarrett2006}.

Another major simplifying assumption in our model was that how individuals themselves were produced (i.e., by selfing or outcrossing) affects the level of inbreeding depression they suffer. In reality, the history of consecutive generations of inbreeding in each individual's lineage will influence the severity of inbreeding depression they experience, particularly when inbreeding depression is caused primarily by recessive deleterious mutations. It would be an interesting and feasible extension of our modeling framework to expand the individual state space to include selfing cohorts (i.e., first generation selfing, second generation selfing, etc.), as in the models of \citet{kelly1999response,kelly2007mutation}, thereby enabling a more biologically plausible approach to modelling self-fertilization.

@@

As described in the main text, we incorporated empirical estimates of the population selfing rate and inbreeding depression for two intensively studied populations in the Cascade Mountains of Oregon into our parameterized model (Iron Mountain and Cone Peak; \citealt{Willis1993,Willis1999a,Willis1999b}). Using the data provided in Table.~2 of \citet{Willis1993}, we calculated the proportional decrease due to inbreeding depression in three fitness components that were also included in vital rate estimates of \citet{PetersonEtAl2016}: seed germination rate ($G$), flower number ($F$), and overwinter survival ($S$). In each case, we calculated the inbreeding depression terms as $\delta_i = 1 - w_{\text{inbred}}/w_{\text{outcross}}$. To incorporate these terms into our evolutionary demographic model, we multiplied each of the three terms $G$, $F$, and $S$ in Eq(\ref{eq:AtildeMimulus}) by a corresponding proportional inbreeding depression term $(1 - \delta_i)$, where $i \in \{ G,F,S\}$. The largest field-estimated selfing rate for this same Iron Mountain population was $C = 0.29$ \citep{Willis1993}, which we use in our calculations for fig.~\ref{fig:mimulusFig}.
